---
title: "JSDM_helminths"
author: "Aimara Planillo"
date: "8/28/2020"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# JSDM model for fox parasites

We are going to analyse the presence of Helminth parasites in foxes detected 
by metabarcoding. 

There are 82 OTU that correspond to 82 Helminth genera


Response: 82 helminth genera
Type of response: Presence/Absence

Explanatory variables:
- noise (dB(a)): buffer 1000m around fox.
- impervious surface (%): buffer 1000m around fox.
- tree cover (%): buffer 1000m around fox.
- arthrhopod index: buffer 1000m around fox.
- sex fox: male, female
- weight fox (kg)


Traits of parasites:
- transmission type
- human related: yes/not affects humans

Randon structure:
- spatial

### Packages
```{r}
library(tidyverse)
library(Hmsc)
library(abind)
library(MCMCvis)
library(corrplot)
library(reshape2)
```


### Workspace
```{r}
Wd <- getwd()

rawdata_wd <- file.path(Wd, "4_raw_data")
procdata_wd <- file.path(Wd, "6_processed_data")

model_wd <- file.path(Wd, "8_results/20200828/Model_output")
results_wd <- file.path(Wd, "8_results/20200828/Model_results")
```

### Load data
```{r}
# Helminth OTU detected in all foxes
helminths <- readRDS(paste0(procdata_wd, "/genus_binary_grouped.rds"))
# Helminth traits
traits <- readRDS(paste0(procdata_wd, "/traits_grouped.rds"))
# Foxes characteristics for Berlin
foxes <- read.csv(paste0(procdata_wd, "/Berlin_foxes_envir_variables_epsg25833.csv"))

head(helminths)
str(helminths)

head(traits)
str(traits)

head(foxes)
str(foxes)

```

For the JSDM we need:
- a matrix of species: sites in rows (foxes) and species in columns (OTU)
- a data frame with the environmental covariates for sites: sites (foxes) in rows
- a trait data frame with the species in rows (OTU), same name and order as in species matrix

## Get species matrix and explanatory variables
```{r}
# traspose OTU data
helminths_t <- t(helminths) %>%
  as.data.frame() %>%
  rownames_to_column(var = "IZW_ID")

nrow(helminths_t)
# [1] 217
nrow(foxes)
# [1] 147

# to be sure that the order and the foxes are the same, we merge response and explanatory tables
# this way we also eliminate samples in the response matrix from foxes not included in the explanatory table
alldata <- merge(foxes, helminths_t, by = "IZW_ID")

summary(alldata)
nrow(alldata)
# [1] 147 foxes
ncol(alldata)

alldata %>%
  select_if(is.numeric) %>%
 colSums

# some columns sum 0. Remove the OTU with no observations from the response
response_data <- alldata%>%
  select(14:95) %>%
  select(which(!colSums(.) %in% 0)) %>%
  as.matrix()

nrow(response_data)
# 147
ncol(response_data)
# 68
colSums(response_data)


# predictors
envcov_cor <- cor(alldata[5:11], use = "pairwise.complete.obs")
# write.csv(envcov_cor, paste0(Wd, "/8_results/20200828/correlations_env_covariates.csv"), row.names = TRUE)

envcov_data <- alldata %>%
  select(IZW_ID, sex, age, tree_cover_1000m, noise_1000m, Human_pop_1000m, Arthrop_1000m)
#weight not included because NA

nrow(envcov_data)


plot(response_data[,30])
```

## Get trait data
```{r}
# get otu names
OTU_genera <- as.data.frame(colnames(response_data))
colnames(OTU_genera) <- "t.genus"

table(traits$transmission)

# simplify trait characteristics
traits_tmp <- traits %>%
  mutate(human.rel = case_when(
    human.related == 0 ~ "No",
    human.related == 1 ~ "Yes",
    TRUE ~ "Unknown")) %>%
  mutate(pet.rel = case_when(
    pet.related == 0 ~ "No",
    pet.related == 1 ~ "Yes",
    TRUE ~ "Unknown")) %>%
  mutate(livestock.rel = case_when(
    livestock.related == 0 ~ "No",
    livestock.related == 1 ~ "Yes",
    TRUE ~ "Unknown")) %>%
  mutate(transmission.3class = case_when(
    transmission == "diet" ~ "diet", 
    is.na(transmission) ~ "Unknown", 
    TRUE ~ "Others"))

table(traits_tmp$human.rel)
table(traits_tmp$pet.rel)

traits_tmp <- merge(OTU_genera, traits_tmp, by = "t.genus")

trait_data <- traits_tmp %>%
  column_to_rownames("t.genus") %>%
  select(human.rel, pet.rel, transmission.3class) 

nrow(trait_data)
```

## get spatial data
```{r}
xyData <- alldata %>%
  transmute(x.coord = coords.x1, y.coord = coords.x2)

```


# Set up the model


### Study design
```{r}
studyDesign <- data.frame(site = as.factor(envcov_data$IZW_ID))
rownames(xyData) <- as.factor(envcov_data$IZW_ID)
rL <- HmscRandomLevel(sData = xyData)
rL$s
```

### Define MCMC parameters
```{r}
thin <- 10
samples <- 10000
transient <- 1000
nChains <- 3
verbose <- 1000

# Regression formula for environmental covariates
XFormula.OTU = ~ sex + 
  tree_cover_1000m + noise_1000m + Human_pop_1000m + Arthrop_1000m
#weight not included because NAs


# Regression formula for traits
TrFormula.OTU = ~ human.rel + pet.rel + transmission.3class
```

## *BINOMIAL DISTRIBUTION* ~> PROBIT MODEL

Untransformed data

## Fit models for PRESENCE/ABSENCE  data 

```{r}
m <- Hmsc(Y = response_data, XData = envcov_data, XFormula = XFormula.OTU,
           studyDesign=studyDesign, ranLevels=list(site=rL),
          TrFormula = TrFormula.OTU, TrData = trait_data,
          distr = "probit")
```

## Run model
```{r}
# Run Markov Chains
# test
m <- sampleMcmc(m, thin = 1, samples = 50)

# the real model
m <- sampleMcmc(m, thin = thin, samples = samples, transient = transient, 
                nChains = nChains, verbose = verbose, nParallel = nChains)

# Set a name for the model
filename <- paste0(model_wd, "/JSDM_metabarcoding_foxes_berlin_20200828.rds")

#Save de model
saveRDS(m,file=filename) 
```


## Model convergence 

We evaluate MCMC convergence in terms of two kinds of parameters that we are especially interested in:
the species niches Beta, influence of traits on species niches Gamma, and the residual species associations Omega. 
The strength of phylogenetic signal rho was not included in this model

Evaluate convergence: Effective sample size and gelman-rubin diagnostic (potencial reduction factor)
```{r model convergence}
mpost <- convertToCodaObject(m)

# Numerical output
ess.beta <- effectiveSize(mpost$Beta)
gd.beta <- gelman.diag(mpost$Beta, multivariate = FALSE)$psrf
ess.gamma <- effectiveSize(mpost$Gamma)
gd.gamma <- gelman.diag(mpost$Gamma, multivariate = FALSE)$psrf
ess.omega <- effectiveSize(mpost$Omega[[1]])
gd.omega <- gelman.diag(mpost$Omega[[1]], multivariate = FALSE)$psrf

convergence.names <- c("ess.beta", "ess.gamma", "ess.omega", 
                       "gd.beta", "gd.gamma", "gd.omega")
convergence.list <- list(ess.beta, ess.gamma, ess.omega, 
                         gd.beta, gd.gamma, gd.omega)
for (i in 1:length(convergence.names)){
  write.csv(convergence.list[[i]], 
            paste0(model_wd, "/", convergence.names[i], ".csv"))
}

# Graphical output
png(paste0(model_wd, "/Metabarcoding_model_20200828_convergence_hist.png"), width = 800, height = 1000,
    pointsize = 20)
par(mfrow=c(3,2))
hist(ess.beta, main = "ess(beta)_Metabarcoding_foxes")
hist(ess.gamma, main = "ess(gamma)_Metabarcoding_foxes")
hist(ess.omega, main = "ess(omega)_Metabarcoding_foxes")
hist(gd.beta, main = "psrf(beta)_Metabarcoding_foxes")
hist(gd.gamma, main = "psrf(gamma)_Metabarcoding_foxes")
hist(gd.omega, main = "psrf(omega)_Metabarcoding_foxes")
dev.off()

# Save plots of the chains
MCMCtrace(mpost$Beta, 
          pdf = TRUE, 
          open_pdf = FALSE,
          filename = "Metabarcoding_model_20200828_MCMCtrace_beta",
          wd = model_wd)
MCMCtrace(mpost$Gamma, 
          pdf = TRUE, 
          open_pdf = FALSE,
          filename = "Metabarcoding_model_20200828_MCMCtrace_gamma",
          wd = model_wd)
MCMCtrace(mpost$Omega[[1]], 
          pdf = TRUE, 
          open_pdf = FALSE,
          filename = "Metabarcoding_model_20200828_MCMCtrace_omega",
          wd = model_wd)

par(mfrow=c(1,1))
```

```{r}
mean(ess.beta)
mean(gd.beta)

mean(ess.gamma)
mean(gd.gamma)

mean(ess.omega)
mean(gd.omega)

```

```{r}
saveRDS(mpost, paste0(model_wd, "/Metabarcoding_model_20200828_mpost_coda.rds"))
mpost <- readRDS(paste0(model_wd, "/Metabarcoding_model_20200828_mpost_coda.rds"))
```


## Model Fit (R2)
```{r model fit}
# Explanatory R2. Get predictions for the observed values 
preds <- computePredictedValues(m, expected = TRUE)
preds.values <- apply(abind(preds,along=3),c(1,2), median) # Median of the predictions
write.csv(preds.values, paste0(model_wd, "/Metabarcoding_model_20200828_Predicted_values_median.csv"))
preds.values.mean <- apply(abind(preds, along = 3), c (1,2), mean) # Mean of the predictions
write.csv(preds.values.mean, paste0(model_wd, "/Metabarcoding_model_20200828_Predicted_values_mean.csv"))

# R2 with the built in function
modelr2.explanatory <- evaluateModelFit(hM = m, predY = preds)
modelr2.explanatory
```


```{r}
# AUC of the model
mean(modelr2.explanatory$AUC)
# [1] 0.8524357

temp.r2 <- cbind.data.frame(r2 = as.numeric(modelr2.explanatory$TjurR2), 
                            species = as.character(m$spNames))
ggplot(data = temp.r2, aes(y=r2, x=species))+
  geom_point()
write.csv(modelr2.explanatory, paste0(model_wd, "/Metabarcoding_model_20200828_explanatory_R2_default.csv"), row.names = FALSE)


# R2 Manually comparing observed vs predicted
R2.sp <- matrix(NA, m$ns, 1)
for (i in 1:m$ns) {
  R2.sp[i, ] <- cor(preds.values.mean[, i],m$Y[, i])^2
}

mean(R2.sp, na.rm=TRUE)
# [1] 

# Plot of the default R2
ggplot(data = as.data.frame(modelr2.explanatory$TjurR2), aes(y=modelr2.explanatory$TjurR2, x=temp.r2$species))+
  geom_point() +
  xlab("Species") +
  ylab("R2 for each species") +
  ggtitle(paste0("Helminth OTU - Default Tjur R2 species \n R2 = ", 
                   round(mean(modelr2.explanatory$TjurR2, na.rm = TRUE), 2)))
ggsave(paste0(model_wd, "/Plot_Metabarcoding_model_20200828_R2_species_default.png"))


# Plot of the manual R2
ggplot(data = as.data.frame(R2.sp), aes(y=V1, x=temp.r2$species))+
  geom_point() +
  xlab("Species") +
  ylab("R2 for each species") +
  ggtitle(paste0("Helminth OTU - Obs vs pred R2 species \n R2 = ", 
                   round(mean(R2.sp, na.rm = TRUE), 2)))
ggsave(paste0(model_wd, "/Plot_Metabarcoding_model_20200828_R2_species_manual.png"))
```

```{r site R2}
R2.site <- matrix(NA, m$ny, 1)
for (i in 1:m$ny) {
  R2.site[i, ] <- cor(preds.values.mean[i, ], m$Y[i, ])^2
}
mean(R2.site, na.rm = TRUE)
# [1] 


ggplot(data = as.data.frame(R2.site), aes(y=V1, x=seq(1, nrow(R2.site), by =1)))+
  geom_point() +
  xlab("Site index") +
  ylab("R2 for each species") +
  ggtitle(paste0("Helminth OTU - Obs vs pred R2 sites \n R2 = ", 
                   round(mean(R2.site, na.rm = TRUE), 2)))
ggsave(paste0(model_wd, "/Plot_Metabarcoding_model_20200828_R2_sites_manual.png"))


# prevalence and r2
ggplot(data = as.data.frame(R2.sp), aes(y=V1, x=colSums(m$Y) / m$ny))+
  geom_point() +
  xlab("Proportion of Sites") +
  ylab("R2 for each species") +
  ggtitle(paste0("Helminth OTU - Explanatory R2: proportion of sites present\n R2 = ", 
                   round(mean(R2.site, na.rm = TRUE), 2)))
ggsave(paste0(model_wd, "/Plot_Metabarcoding_model_20200828_R2_prevalence_manual.png"))


# compare predicted and observed col sums (one data point is a species)
png(paste0(det_model_wd, '/Plot_Metabarcoding_model_20200828_Obs_vs_Pred.png'))
plot(colSums(preds.values)~colSums(m$Y),
     main = paste0("Helminth OTU", " - Observed vs Predicted Species presences \n Mean R2 = ", round(mean(R2.sp, na.rm = TRUE), 2)),
     xlab = "n observed sites",
     ylab = "n predicted sites",
     pch = 16)
abline(0,1, col = "red", 
       lty = 4,
       lwd = 1)
dev.off()
```

## Beta values

```{r betas}
Beta.results <- as.data.frame(MCMCsummary(mpost$Beta))
write.csv(Beta.results, paste0(model_wd, "/Metabarcoding_model_20200828_beta_coeffients.csv"), row.names = TRUE)
# Default beta plot in Hmsc package
postBeta <- getPostEstimate(m, parName = "Beta")
saveRDS(postBeta, file=paste0(model_wd, "/Metabarcoding_model_20200828_postBeta.rds"))
png(paste0(results_wd, "/Plot_Metabarcoding_model_20200828_beta_support95.png"))
plotBeta(m, post = postBeta, param = "Support", supportLevel = 0.95)
dev.off()
  
png(paste0(results_wd, "/Plot_Metabarcoding_model_20200828_beta_support75.png"))
plotBeta(m, post = postBeta, param = "Support", supportLevel = 0.75)
dev.off()

# my plots
# Coef plot for betas for each species
pdf(paste0(results_wd, "/Plot_Metabarcoding_model_20200828_beta_coef_plots.pdf"))
MCMCplot(mpost$Beta, 
           ref_ovl = TRUE,
           rank = T,
           xlab = 'ESTIMATE',
           sz_labels = 0.3,
           sz_med = 1,
           sz_thick = 3,
           sz_thin = 1,
           sz_ax = 1,
           sz_main_txt = 1)
dev.off()


# Print a plot for each predictor
n.cov <- length(m$covNames) # Number of covariates without the intercept
  var.code <- vector()
  for (i in 1:n.cov){
    var.code[i] <- paste0("C", i)
  }
  
var.name <- as.vector(m$covNames[1:n.cov])
predictors <- as.data.frame(cbind(var.code, var.name))
  
for (i in 1:nrow(predictors)){
    pdf(paste0(results_wd, "/Betas_covariates_coef_plot_", 
               var.name[i], "_Metabarcoding_model_20200828.pdf"))
    MCMCplot(mpost$Beta,
             params = predictors[i,1],
             ISB = FALSE,
             ref_ovl = TRUE,
             rank = FALSE,
             xlab = 'ESTIMATE',
             main = predictors[i,2],
             sz_labels = 0.5,
             sz_med = 1,
             sz_thick = 1,
             sz_thin = 1,
             sz_ax = 1,
             sz_main_txt = 1)
    dev.off()
}

```


## Gamma values
```{r}
Gamma.results <- as.data.frame(MCMCsummary(mpost$Gamma))
write.csv(Gamma.results, paste0(model_wd, "/Metabarcoding_model_20200828_gamma_coeffients.csv"), 
            row.names = FALSE)
# Default gamma plot in Hmsc package
postGamma <- getPostEstimate(m, parName = "Gamma")
png(paste0(results_wd, "/Metabarcoding_model_20200828_default_gammaplot095.png"))
plotGamma(hM = m, post = postGamma, param = "Support", supportLevel = 0.95)
dev.off()

png(paste0(results_wd, "/Metabarcoding_model_20200828_default_gammaplot075.png"))
plotGamma(hM = m, post = postGamma, param = "Support", supportLevel = 0.75)
dev.off()
  
# Coef. gammma
pdf(paste0(results_wd, "/Metabarcoding_model_20200828_gamma_coef_plots.pdf"))
MCMCplot(mpost$Gamma, ref_ovl = TRUE)
dev.off()
```


## Speceis co-occurrences 

We do it throuch species associations in residual random variance

```{r sp associations}
OmegaCor <- computeAssociations(m)
class(OmegaCor)
saveRDS(OmegaCor, file=paste0(model_wd, "/Metabarcoding_model_20200828_OmegaCor.rds"))

OmegaCor[[1]]$mean
OmegaCor[[1]]$support

# Default plot in Hmsc package
supportLevel <- 0.95

toPlot <- ((OmegaCor[[1]]$support > supportLevel)
           + (OmegaCor[[1]]$support < (1 - supportLevel)) > 0) * OmegaCor[[1]]$mean
png(paste0(results_wd, "/Metabarcoding_model_20200828_default_omegaplot95.png"))
corrplot(toPlot, method = "color", 
         col = colorRampPalette(c("blue", "white", "red"))(200),
         title = paste0("random effect level: ", m$rLNames[1]), 
         mar = c(0,0,1,0))
dev.off()

supportLevel <- 0.75
toPlot <- ((OmegaCor[[1]]$support > supportLevel)
           + (OmegaCor[[1]]$support < (1 - supportLevel)) > 0) * OmegaCor[[1]]$mean
png(paste0(results_wd, "/Metabarcoding_model_20200828_default_omegaplot75.png"))
corrplot(toPlot, method = "color", 
         col = colorRampPalette(c("blue", "white", "red"))(200),
         title = paste0("random effect level: ", m$rLNames[1]), 
         mar = c(0,0,1,0))
dev.off()

par(mar = c (5,5,4,5))
```


```{r save associations}
assoc.mean <- melt(OmegaCor[[1]]$mean)
assoc.support <- melt(OmegaCor[[1]]$support)

nrow(assoc.mean); nrow(assoc.support)

associations <- cbind.data.frame(assoc.mean, support = assoc.support$value)
colnames(associations) <- c("species1", "species2", "mean", "support")

associations

write.csv(associations, paste0(model_wd, "/Metabarcoding_model_20200828_sp_associations.csv"), row.names = FALSE)
```

## Variance partitioning

The order of the variables, if they are continuous, is 
1. intercept(this can be in any group)
2. first variable
3. second variable
ETC.

The formulas we used for running the models area: 
XFormula = ~ (tree.100m + open.green100m + noise.100m + pop.100m) * prey.abu

```{r variance partitioning}
# design matrix
head(m$X)

# Total variance explained by hte model can be partition into the contributions
# of each fixed effect (or group) and random effect
# intercept can go in any group. Here in the first to simplify
# Groups: season, garden, landscape, cats
VP <- computeVariancePartitioning(m, group = c(1,1, 2, 3,3, 2), groupnames = c("sex", "natural", "urban"))

# Save the Variance partitioning info
VP$vals
VP$R2T

saveRDS(VP, file = paste0(model_wd, "/Metabarcoding_model_20200828_varpart.rds"))

VP.table <- as.data.frame(VP$vals) 
VP.table
write.csv(VP.table, paste0(model_wd, "/Metabarcoding_model_20200828_Varpart_values.csv"))


# plot var part
png(paste0(results_wd, "/Plot_Metabarcoding_model_20200828_default_VP.png"), 
    width = 800)
plotVariancePartitioning(m, VP = VP, las = 2, cex.names = 0.8)
title(main = "\n \nDETECTION Helminth genera")
dev.off()
```



















